---
title: "Lab 1"
author: "Eric, Samir, Carlos"
date: "September 24, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Loading


```{r}

# Load libraries
library(readr)
library(car)
require(dplyr)

# Load data from Csv
public.opinion <- data.frame(read_csv("public_opinion.csv"))

# Turn categorical / binary features to factors
public.opinion$sanders_preference <- factor(public.opinion$sanders_preference)
public.opinion$party <- factor(public.opinion$party)
public.opinion$race_white <- factor(public.opinion$race_white)
public.opinion$gender <- factor(public.opinion$gender)

# Show the first few rows
head(public.opinion)


```


# Data Cleasing

Below we verify limits, missing values and potential inconsistencies in the data, prior to diving into data analysis.

## Data Cleansing: Party

```{r}

levels(public.opinion$party)
describe(public.opinion$party)
```

Text: 3 levels, no missing values. 38.4% independent not in any party. 

## Data Cleansing: Gender


```{r}

levels(public.opinion$gender)
describe(public.opinion$gender)
```

## Data Cleansing: Race white

```{r}

levels(public.opinion$race_white)
describe(public.opinion$race_white)
```

## Data Cleansing: Birth year

```{r}

describe(public.opinion$birthyr)

ggplot(public.opinion, aes(x = birthyr)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, fill="#0072B2", colour="black") +
  ggtitle("Birth Year") + 
  theme(plot.title = element_text(lineheight=1, face="bold"))
```

## Data Cleansing: Sanders Preference

```{r}

levels(public.opinion$race_white)
describe(public.opinion$race_white)
```

# Data Analysis

## Data Analysis: Birth year

```{r}

ggplot(public.opinion, aes(sanders_preference, birthyr)) +
  geom_boxplot(aes(fill = sanders_preference)) + 
  geom_jitter() +
  ggtitle("Birth year by Sanders Preference") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 


```


```{r}

ggplot(public.opinion, aes(birthyr, fill = sanders_preference, colour = sanders_preference)) +
  geom_density(alpha=0.2)

```

## Data Analysis: Party

```{r}

# TODO: Here plot / table of sanders preference by party and demonstrate any interesting points


```

```{r}
ggplot(public.opinion, aes(birthyr, fill = party, colour = party)) +
  geom_density(alpha=0.2)

```

## Data Analysis: Gender

```{r}

```
## Data Analysis: Race White


```{r}

# TODO: more work here but not sure if gender will be useful. Support with data if it wont.
```


# Model Selection

# Test Model Assumptions

# Model Assessment
