---
title: "Lab 1"
author: "Eric, Samir, Carlos"
date: "September 24, 2017"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data Loading


```{r}

# Load libraries
library(readr)
library(car)
require(dplyr)

# Load data from Csv
public.opinion <- data.frame(read_csv("public_opinion.csv"))

# Turn categorical / binary features to factors
public.opinion$sanders_preference <- factor(public.opinion$sanders_preference)
public.opinion$party <- factor(public.opinion$party)
public.opinion$race_white <- factor(public.opinion$race_white)
public.opinion$gender <- factor(public.opinion$gender)

# Recode gender
public.opinion$gender.name[public.opinion$gender == 1] <- "Male"
public.opinion$gender.name[public.opinion$gender == 2] <- "Female"
public.opinion$gender.name = factor(public.opinion$gender.name)

# Recode birth year into age
public.opinion$age <- 2017 - public.opinion$birthyr

# Recode party
public.opinion$party.name[public.opinion$party == 1] <- "Democrat"
public.opinion$party.name[public.opinion$party == 2] <- "Independent"
public.opinion$party.name[public.opinion$party == 3] <- "Republican"
public.opinion$party.name = factor(public.opinion$party.name)

# Remove records without sanders preference as it is our outcome variable, and without it we can't use those records
# Actually move this below the data cleansing or unify
public.opinion <- public.opinion[complete.cases(public.opinion), ]

summary(public.opinion[!is.na(public.opinion$sanders_preference)])

# Show the first few rows
head(public.opinion)

str(public.opinion)

```


# Data Cleasing

Below we verify limits, missing values and potential inconsistencies in the data, prior to diving into data analysis.

## Data Cleansing: Party

```{r}

levels(public.opinion$party)
describe(public.opinion$party)
```

Text: 3 levels, no missing values. 38.4% independent not in any party. 

## Data Cleansing: Gender


```{r}

levels(public.opinion$gender)
describe(public.opinion$gender)
```

## Data Cleansing: Race white

```{r}

levels(public.opinion$race_white)
describe(public.opinion$race_white)
```

## Data Cleansing: Birth year

```{r}

describe(public.opinion$birthyr)

ggplot(public.opinion, aes(x = birthyr)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, fill="#0072B2", colour="black") +
  ggtitle("Birth Year") + 
  theme(plot.title = element_text(lineheight=1, face="bold"))
```

## Data Cleansing: Sanders Preference

```{r}

levels(public.opinion$race_white)
describe(public.opinion$race_white)
```

# Data Analysis

## Data Analysis: Age

```{r}

ggplot(public.opinion, aes(sanders_preference, age)) +
  geom_boxplot(aes(fill = sanders_preference)) + 
  geom_jitter() +
  ggtitle("Age by Sanders Preference") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 


```


```{r}

ggplot(public.opinion, aes(age, fill = sanders_preference, colour = sanders_preference)) +
  geom_density(alpha=0.2)

```

## Data Analysis: Party

```{r}

# TODO: Here plot / table of sanders preference by party and demonstrate any interesting points


```

```{r}

ggplot(public.opinion, aes(party.name, age)) +
  geom_boxplot(aes(fill = party.name)) + 
  #geom_jitter() +
  ggtitle("Age vs party segregated on sanders preference") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 


```


```{r}

ggplot(public.opinion, aes(party.name, age)) +
  geom_boxplot(aes(fill = sanders_preference)) + 
  #geom_jitter() +
  ggtitle("Age vs party segregated on sanders preference") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 


```

```{r}
ggplot(public.opinion, aes(age, fill = party, colour = party.name)) +
  geom_density(alpha=0.2)

```

## Data Analysis: Gender

```{r}
table(public.opinion$sanders_preference, public.opinion$gender.name)
```
## Data Analysis: Race White

```{r}

ggplot(public.opinion, aes(race_white, age)) +
  geom_boxplot(aes(fill = sanders_preference)) + 
  #geom_jitter() +
  ggtitle("Age vs party segregated on sanders preference") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 


```

```{r}
ggplot(public.opinion, aes(gender.name, age)) +
  geom_boxplot(aes(fill = gender.name)) + 
  geom_jitter() +
  ggtitle("Age vs Race") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 
```

```{r}

ggplot(public.opinion, aes(race_white, age)) +
  geom_boxplot(aes(fill = race_white)) + 
  #geom_jitter() +
  ggtitle("Age vs Race") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 

# Bias in the ages for the sample!!!!!

```

# Model Selection

```{r}
model <- glm(sanders_preference ~ age + gender.name + race_white + party.name + age:race_white, family =  binomial(link = 'logit'), data = public.opinion)
summary(model)
```


```{r}
model.simple <- glm(sanders_preference ~ age + race_white + party.name, family =  binomial(link = 'logit'), data = public.opinion)
summary(model.simple)
```

# Model

$$logit(\hat{\pi}) = -0.114017 - 0.01248 age + 0.872782 race_white1 + 0.713501 party.nameIndependent + 0.594231 party.nameRepublican$$

# Assessment

## OR for level Party Independent

```{r}
OR_independent = exp(0.713501)
OR_independent
```

The odds of supporting sanders are 2.04 times as large for for individuals of independent political party compared to democrats where the age and race are held constant.


## OR for level Party Republican

```{r}
OR_republican = exp(0.594231)
OR_republican
```

The odds of supporting sanders are 1.81 times as large for for individuals of independent political party compared to democrats where the age and race are held constant.

## OR for level white race

```{r}
OR_white = exp(0.872782)
OR_white
```

The odds of supporting sanders are 2.39 times as large for for individuals of white color compared to non-white where the age and political party are held constant.

# Hypothesis testing

Our reduced model

$$logit(\hat{\pi}) = -0.114017 - 0.01248 age + 0.872782 race_white1 + 0.713501 party.nameIndependent + 0.594231 party.nameRepublican$$

# Likelihood ratio test for full model

```{r}
Anova(mod = model, test = "LR")
```

# Likelihood ratio test for reduced model

```{r}
Anova(mod = model, test = "LR")
```


```{r}
w <- aggregate(formula = sanders_preference ~ age + race_white + party.name, data = public.opinion, FUN = sum)

n <- aggregate(formula = sanders_preference ~ age + race_white + party.name, data = public.opinion, FUN = length)

w.n <- data.frame(age = w$age, success = w$sanders_preference, trials = n$sanders_preference, proportion = round(w$sanders_preference/n$sanders_preference, 4))
head(w.n)

win.graph(width = 7, height = 6, pointsize = 12)
plot(x = w$age, y = )

```

