---
title: "Statistical Methods for Discrete Response, Time Series, and Panel Data (W271): Lab 1"
author: "Eric Yang, Samir Datta, Carlos Castro"
date: "September 30, 2017"
output: pdf_document
---

# Data loading

Now we load the data and transform the factor variables for more semantically meaningful levels.

```{r}
library(ggplot2)
theme_set(theme_bw())
library(car)
publicopinion <- read.csv('public_opinion.csv')


#create subset without issing values
publicopinion_narm <- publicopinion[!is.na(publicopinion$sanders_preference),]
#create age variable
publicopinion$age <- 2017 - publicopinion$birthyr
#turn variables into factors
publicopinion$partyfactor <- ifelse(publicopinion$party==1, 'Democrat', 
                                    ifelse(publicopinion$party==2, 'Other', 
                                           'Republican'))


publicopinion$genderfactor <- ifelse(publicopinion$gender==1,'Male', 'Female')
publicopinion$racefactor <- ifelse(publicopinion$race_white==1, 'White', 'Non-White')
publicopinion$spfactor <- ifelse(publicopinion$sanders_preference==1, "Yes", "No")

```

# 1.a Model

## Model Overview
The model looks at the likelihood of supporting bernie sanders based on 4 explanatory variables. The three explantory variables are age as of 2017, race_white (non white vs. white), independent voter (baseline democrat), republican (baseline democrat).  

## Age
The age variable was caculated as of 2017 based off the birth year variable in the original dataset. Age has a negative coefficient which is statistically significant showing that an increase in age reduces the likelihood of supporting sanders.  The odds_age object shows that for every 10 years decrease in age we see a ~1.13 increase in the odds of supporting sanders. 

## Racefactor
Racefactor = 0 is non white, racefactor = 1 is white. Racefator has a positive coefficient in our model that is statistically significant.  Looking at the odds ratios we see that being white increases odds of supporting sanders by 2.39 holding all other variables constant.  

## Partyfactor
Both the independent and republican variables have statistically significant positive coefficients.  This shows that either party affiliation versus democrat would increase the probability of supporting sanders.  A look at the odds ratio shows an odds increase of 2.04 for independents and 1.81 for republicans compared to democrats. For independents, the increase in odds of preferring Sanders is 1.13 when compared to republicans. 

```{r}
#Section 1a.
model.final <- glm(sanders_preference ~ age + racefactor + partyfactor, data = publicopinion, family = binomial(link = logit))

summary(model.final)

odds_age <- exp(model.final$coefficients[2]*-10)
odds_age

odds_white<- exp(model.final$coefficients[3])
odds_white

odds_ind <- exp(model.final$coefficients[4])
odds_ind

odds_rep <- exp(model.final$coefficients[5])
odds_rep
```
# 1.b EDA
## Overview
```{r}
summary(publicopinion)
```
```{r}
publicopinion[is.na(publicopinion$sanders_preference),]
```

There are nine observations missing a value for the variable of interest sanders_preference. There does not appear to be any pattern to these missing values, which would suggest they are missing at random and not indicative of any sampling bias. We used this data ignoring these nine missing values which left us with a complete dataset. 

## Age
```{r}
# Distribution of Age
ggplot(publicopinion, aes(x = age)) +
  geom_histogram(aes(y = ..density..), binwidth = 0.2, fill="#0072B2", colour="black") +
  ggtitle("Age") + 
  theme(plot.title = element_text(lineheight=1, face="bold"))
```
Ages range from 20 to 96.  Appears to have a bimodal distribution with peaks in the mid 60s and mid 40s and a dip in the 50s. 

```{r}

ggplot(publicopinion, aes(partyfactor, age)) +
  geom_boxplot(aes(fill = spfactor)) + 
  #geom_jitter() +
  ggtitle("Age vs party segregated on sanders preference") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 

```
The box plot above shows a difference in age between party affiliation. Republican voters skew older for both sanders preference outcomes with the median age in the mid fifties.  The indpendent voters skew younger compared to republicans with median age in the mid 40s.  There appears to be the largest gap between democratic voters when taking into account sanders preference.  Older democratic voters appear less likely to support sanders than younger demcratic voters.

```{r}
ggp <- ggplot(publicopinion_narm, aes(x=age, y=sanders_preference))

ggp + geom_point()+
  geom_smooth(method="loess", se=T)+
  ylab("Percent with Sanders preference")+
  xlab("Age")+
  ggtitle("Sanders preference grouped by age")+
  theme(plot.title=element_text(hjust=.5))
```

Above is a scatterplot of all of the points in the set, with age on the x-axis and the binary variable sanders_preference on the y-axis. Displaying the dots like this is not necessarily informative, but the loess smooth curve - and standard error ribbon - reveals an interesting trend. Below the age 50, there seems to be a trend for younger voters to prefer Sanders. However, this trend is also seen in the opposite direction for voters above around 70. This would suggest that including a quadratic term for age might be useful. However, it is important to note that the standard error ribbon is very large towards the older end of the age range, which is indicative of how few voters of that age range we really have. While we should try modeling a quadratic term for age, we should be careful not to over-interpret any result based off insufficient data.


## Party
```{r}
ggplot(publicopinion, aes(partyfactor, age)) +
  geom_boxplot(aes(fill = partyfactor)) + 
  geom_jitter() +
  ggtitle("Party Affiliation by Age") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 
```
Democrats median age is 49, Independents are 45, Republicans are 54.  Republicans have a higher age range and have more observations in the 80+ age range.  We should add party to control for party affiliation since there seems to be a difference in age between groups. 

```{r}
po_party_agg <- with(publicopinion_narm,
                     aggregate(cbind(100*sanders_preference),
                               list(partyfactor=partyfactor),
                               mean))
po_party_agg$n <- with(publicopinion_narm,
                     aggregate(cbind(100*sanders_preference),
                               list(partyfactor=partyfactor),
                               length))[,2]

po_party_df <- data.frame(sanders_pref_percent=po_party_agg$V1,
                          party=po_party_agg$partyfactor, 
                          sample_percent = 100*po_party_agg$n/1191)
po_party_df
```

Republicans, who make up 23% of the sample, are slightly underrepresented compared to Democrats and Other, but not to the extent that we should be worried about sampling bias. A slight majority (55%) of Democrats polled preferred Clinton to Sanders, while a majority of Republicans (64%) and Other (66%) preferred Sanders. This difference supports including party as an explanatory variable.
```{r}
age_bin_agg_party <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       party=partyfactor), mean))
age_bin_agg_party$n <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       party=partyfactor), length))[,3]

ggp <- ggplot(age_bin_agg_party, aes(x=agebin, y=sanders_preference,
                                     color=party, size=n))

ggp + geom_point(aes(color=party))+
  geom_smooth(method="loess", se=F)+
  ylab("Percent with Sanders preference")+
  xlab("Age")+
  ggtitle("Sanders preference grouped by age\n(divided by party)")+
  theme(plot.title=element_text(hjust=.5))
```
Above is a scatterplot of age on the x-axis, where color represents the political party. Each dot's position on the y-axis represents the percent of people of that specific party and age that preferred Sanders. 

The democrats seem to have a fairly clear relationship with age in that younger democrats look more likely to support Sanders than older ones. The relationship within Republicans is less clear, and for independents, it looks almost quadratic (as the smooth curve lifts upwards both for younger and older voters). The curves are loess smoothed curves and not meant to be a perfect representation of overall trends. However, there is still enough evidence to support at least trying to model an age by party interaction, since it looks like different parties may have different relationships with age.
## Race
```{r}
ggplot(publicopinion, aes(racefactor, age)) +
  geom_boxplot(aes(fill = spfactor)) + 
  #geom_jitter() +
  ggtitle("Age vs party segregated on sanders preference") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 
```
Looking at the race_white by age chart we see the distribution of age with respect to race.  Whites skew older with a median age of 53 compared to that of 43 for non whites.  This large difference in preference suggests we should add the race_white variable so we can control for the effect of race when evaluating age.  Potentially would be interesting to look add an iteraction term the model, age:race_white, to see the effect of age with respect to age and sanders preference.

```{r}
po_race_agg <- with(publicopinion_narm,
                     aggregate(cbind(100*sanders_preference),
                               list(racefactor=racefactor),
                               mean))
po_race_agg$n <- with(publicopinion_narm,
                     aggregate(cbind(100*sanders_preference),
                               list(racefactor=racefactor),
                               length))[,2]

po_race_df <- data.frame(sanders_pref_percent=po_race_agg$V1,
                          party=po_race_agg$racefactor, 
                          sample_percent = 100*po_race_agg$n/1191)
po_race_df
```
White voters make up about 73% of the sample, which is close to the estimated percentage of White people in America, which further supports our sample being representative. There is a large difference between how many white voters (64%) vs. non-whie voters (41%) prefer Sanders.

```{r}
age_bin_agg_race <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       race=racefactor), mean))
age_bin_agg_race$n <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       race=racefactor), length))[,3]

ggp <- ggplot(age_bin_agg_race, aes(x=agebin, y=sanders_preference,
                                     color=race, size=n))

ggp + geom_point(aes(color=race))+
  geom_smooth(method="loess", se=F)+
  ylab("Percent with Sanders preference")+
  xlab("Age")+
  ggtitle("Sanders preference grouped by age\n(divided by race)")+
  theme(plot.title=element_text(hjust=.5))
```
At first glance the relationship seems very different for non-white voters, but this may be an artifact of the loess smoothing curve attempting to compensate for the data points in the youngest age groups. When looking at the overall distribution of the dots it seems that both white and non-white voters have a negative relationship with age.

## Gender
```{r}
ggplot(publicopinion, aes(genderfactor, age)) +
  geom_boxplot(aes(fill = genderfactor)) + 
  geom_jitter() +
  ggtitle("Gender by Age") + 
  theme(plot.title = element_text(lineheight=1, face="bold")) 
```
In the dataset women skew towards an older age, median 52 versus 47 for males.  

```{r}
po_gender_agg <- with(publicopinion_narm,
                     aggregate(cbind(100*sanders_preference),
                               list(genderfactor=genderfactor),
                               mean))
po_gender_agg$n <- with(publicopinion_narm,
                     aggregate(cbind(100*sanders_preference),
                               list(genderfactor=genderfactor),
                               length))[,2]

po_gender_df <- data.frame(sanders_pref_percent=po_gender_agg$V1,
                          party=po_gender_agg$genderfactor, 
                          sample_percent = 100*po_gender_agg$n/1191)
po_gender_df
```

Male and female voters are close to equally represented, both around 50%. Across the sample male and female voters prefer Sanders at almost the same rate (57.5% for male, 57.7% for female)
```{r}
age_bin_agg_gender <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       gender=genderfactor), mean))
age_bin_agg_gender$n <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       gender=genderfactor), length))[,3]

ggp <- ggplot(age_bin_agg_gender, aes(x=agebin, y=sanders_preference,
                                     color=gender, size=n))

ggp + geom_point(aes(color=gender))+
  geom_smooth(method="loess", se=F)+
  ylab("Percent with Sanders preference")+
  xlab("Age")+
  ggtitle("Sanders preference grouped by age\n(divided by gender)")+
  theme(plot.title=element_text(hjust=.5))
```
Towards the younger end of the age range, it seems like the negative relationship between sanders preference and age exists mostly for female voters and not so much for male voters. This suggests that investigating a gender by age interaction may be useful.

# 1.c Alternate models

In our EDA, we determined that race and party had large effects on Sanders preference and would be important to control for. Gender did not seem like it had much explanatory power on its own, although a gender by age interaction seemed plausible. A race by age interacton also looked to be worth testing. Finally, we wanted to test the plausability of a quadratic age term.

###Gender by age interaction

```{r}
glm.out.base <- glm(sanders_preference ~ age + partyfactor + racefactor + 
                 genderfactor, data=publicopinion_narm,
                 family=binomial(link='logit'))
glm.out.int <- glm(sanders_preference ~ age + partyfactor + racefactor + 
                 genderfactor + age:genderfactor, data=publicopinion_narm,
                 family=binomial(link='logit'))

summary(glm.out.base)
```
```{r}
summary(glm.out.int)
```

```{r}
anova(glm.out.base, glm.out.int, test="LR")
```

We noticed that the statistical significance of the age term got slightly stronger in the interaction model than in the base model. In the base model, the coefficient for age represented the relationship between age and Sanders preference across the sampl. In the interaction model, it represents the relationship for the base level of gender only, which is female. The coefficient of the interaction term is .011, which means the coefficient for age in males only would be -.018 + .011 = -.007. We can interpret this to mean that the model is suggesting there is a strong negative relationship for female voters but not male voters.

We have decided against this model for a few reasons. First of all, we know from looking at the data and at both models that the effect of gender is very small. Second of all, the p-values given to the interaction term and the likelihood ratio test are not statistically significant even at p < 0.1. This tells us that although there may be a hint of a gender interaction term it isn't strong enough to justify including it in our final model.

###Party by age interaction

```{r}
glm.out.base <- glm(sanders_preference ~ age + partyfactor + racefactor + 
                 genderfactor, data=publicopinion_narm,
                 family=binomial(link='logit'))
glm.out.int <- glm(sanders_preference ~ age + partyfactor + racefactor + 
                 genderfactor + age:partyfactor, data=publicopinion_narm,
                 family=binomial(link='logit'))

summary(glm.out.base)
```
```{r}
summary(glm.out.int)
```

There are a number of reasons that modeling an age by party interaction does not seem like a good idea. First of all, neither of the interacton terms (age:Other, age:Republican) have large effects. Their coefficients are relatively small compared to the original coefficient of the age term, and their p-values are nowhere near statistical significance (p > .33). The AIC for the model with the interaction term is larger than for the model without it, suggesting that our additional model complexity is not helping the overall model.

```{r}
anova(glm.out.base, glm.out.int, test="LR")
```

The likelihood ratio test, with a p-value of .53, also suggests our interaction model is not more useful than the simpler model. For these reasons we decided to not model an age by party interaction.

###Quadratic age term

```{r}
glm.out.base <- glm(sanders_preference ~ age + partyfactor + racefactor + 
                 genderfactor, data=publicopinion_narm,
                 family=binomial(link='logit'))
glm.out.quad <- glm(sanders_preference ~ age + partyfactor + racefactor + 
                 genderfactor + I(age^2), data=publicopinion_narm,
                 family=binomial(link='logit'))

summary(glm.out.base)
```

```{r}
summary(glm.out.quad)
```

```{r}
anova(glm.out.base, glm.out.quad, test="LR")
```

Like the EDA showed, the model is showing that a quadratic age term might be plausible. The significance of the term in the model is slightly larger than .05 (.06), as is the signifiance of the likelihood ratio test (.06). The AIC of the model with the quadratic term is also lower.

However, we decided to not to include the quadratic age term in the end. Aside from the lack of statistical significance - although it is close - the main reason for this is the lack of representation in the older age range that is driving this result. We would not feel comfortable recommending this model and suggesting that older voters be targeted when we have so few older people that are contributing to this trend.

# 1.d Selected model results

# 1.e Statistical Tests

In the subsections below we perform two statistical tests on our model: Wald test and Likelihood Ratio Test. 

## Wald Test

The summary of our model actually displays the statistics for the Wald Test. Recall that the Wald statistic is given by

$$Z_0 = \frac{\hat{\beta_r - \beta_r}}{\sqrt{Var(\hat{\beta_r})}}$$

We use this statistic to test the null hypothesis $H_0: \beta_r = 0$ versus the alternat hypothesis $H_a: \beta_r \ne 0$. 

For large samples, this test statistic has an approximate standard normal distribution if the null hypothesis is true, and we reject the null hypothesis if the statistic value is unexpected for a standard normal distribution.

Now, since the Wald test statistic is provided automatically for each individual $\beta$ parameter, we summarize the model again with the purpose of analyzing the Wald statistics.

```{r}
summary(model.final)
```

We can observe that for each explanatory variable there is a hypothesis test, with its associated p-value. 

For example, for the $age$ explanatory variable, we observe that the p-value is $0.000664$, which is highly statistically significant, so we can reject the null hypothesis $H_0$ that $\beta_{age} = 0$, which can be interpreted as there is sufficient statistical evidence to indicate that age has an effect on the probability of voters supporting Sanders. For the other explanatory variables we analogously reject the null hypotheses that their corresponding $\beta$ equal $0$, since all the statistics are highly statistically significant.

## Likelihood Ratio Test

We now perform likelihood ratio tests on our model. The LRT statistic is defined as

$$\Lambda = \frac{Maximum\ of\ likelihood\ function\ under\ H_0}{Maximum\ of\ likelihood\ function\ under\ H_0\ or\ H_a}$$

The test, as with the Wald statistic, is for $H_0: \beta_r = 0$ versus $H_a: \beta_r \ne 0$.

We then calculate $-2log(\Lambda)$, and if the null hypothesis is true, then $-2log(\Lambda)$ has an approximate $\chi^2_1$ distribution for a large sample.

In R, we can perform the likelihood ratio test using the Anova function authored by Professor John Fox as part of the Car package. Below are the results of our test.

```{r}
Anova(model.final)
```

For example, for the $age$ explanatory variable, we obtain a statistically significant p-value 0.0006217, rejecting the null hypothesis that $\beta_{age} = 0$. For Race and Party we also obtained statistically significant p-values and we can also reject their null hypothesis, which means that there is evidence that each explanatory variable has an effect on the probability of voters supporting Sanders over Clinton.

## 1.f Age Interpretation and Odds Ratios

### Odds Ratios

First, let's obtain an expression for the effect in the odds of supporting Sanders caused by a $c$ year change in age. For this, we calculate the odds ratio:

$$OR = \frac{Odds_{age+c}}{Odds_{age}} = \frac{e^{\beta_0 + \beta_{age} (age + c) + \beta_{race.white} race.white + \beta_{party.republican} party.republican + \beta_{party.independent} party.independent}}{e^{\beta_0 + \beta_{age} age + \beta_{race.white} race.white + \beta_{party.republican} party.republican + \beta_{party.independent} party.independent}}$$

Which using the properties of exponentiation can be simplified to:

$$OR = e^{c \beta_{age}}$$

For example, we can calculate the odds ratio for a $10$ year decrease in age by inverting the formula:

```{r}
1 / exp(model.final$coefficients[2] * 10)
```

This odds ratio can be interpreted as: the odds of supporting sanders are $1.132925$ times larger for every 10 year decrease in age of the voters.

### Odds Ratios and Confidence Intervals

To include confidence intervals in our odd ratios, we use the Wald confidence interval, which comes from the following expression:

$$c\hat{\beta_{age}} \pm cZ_{1-\alpha/2} \sqrt(Var(\hat{\beta_{age}}))$$

This means we need the variance of $\beta_{age}$, which we can obtain from the variance-covariance matrix for our model:

```{r}
vcov(model.final)
```

Where $var(\hat{\beta_{age}})$ is in the diagonal, with value $1.344168e-05$.

Now we can calculate the intervals:

```{r}
c = 10
alpha = 0.05
confint = c * qnorm(1 - alpha/2) * sqrt(1.344168e-05)

odds.ratio = 1 / exp(model.final$coefficients[2] * 10)

lower = odds.ratio - confint
upper = odds.ratio + confint

odds.ratio
lower
upper
```

So going back to our interpretation, the odds of supporting sanders are $1.132925$ times larger for every 10 year decrease in age of the voters. This ratio, for the 95% confidence interval, can be found between $1.204783$ and $1.061067$.


# 2 Plot: Age vs Predicted probability of supporting Sanders

```{r}
#create variable a to represent ages 20 to 100
#create different y values for each subgroup of party and race - 6 total
#using the model predited probabilities for sanders_preference
a = c(20:100)
y_dem_nw = exp(model.final$coefficients[1] + model.final$coefficients[2]*a)/
       (1+exp(model.final$coefficients[1] + model.final$coefficients[2]*a))

y_oth_nw = exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[3])/
       (1+exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[3]))

y_rep_nw = exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[4])/
       (1+exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[4]))



y_dem_w = exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                model.final$coefficients[5])/
       (1+exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                model.final$coefficients[5]))

y_oth_w = exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[3] +
                model.final$coefficients[5])/
       (1+exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[3] +
                model.final$coefficients[5]))

y_rep_w = exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[4] +
                model.final$coefficients[5])/
       (1+exp(model.final$coefficients[1] + model.final$coefficients[2]*a +
                 model.final$coefficients[4] +
                model.final$coefficients[5]))
        

```

```{r}
age_bin_agg_all <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       party=partyfactor,
                                       race=racefactor), mean))
age_bin_agg_all$n <- with(publicopinion_narm,
                        aggregate(cbind(sanders_preference),
                                  list(agebin=age, 
                                       party=partyfactor,
                                       race=racefactor), length))[,4]
```

```{r}
#plot data and predictions for white voters
ggp <- ggplot(age_bin_agg_all[age_bin_agg_all$race=="White",], aes(x=agebin, y=sanders_preference,
                                     color=party, size=n))

ggp + geom_point()+
  geom_path(inherit.aes=F, data=data.frame(x=a, y=y_dem_w),
             aes(x = a, y = y), color="blue")+
  geom_path(inherit.aes=F, data=data.frame(x=a, y=y_oth_w),
             aes(x = a, y = y), color="green")+
    geom_path(inherit.aes=F, data=data.frame(x=a, y=y_rep_w),
             aes(x = a, y = y), color="red")+
  #facet_grid(party~.)+
  scale_color_manual(values=c("blue", "green", "red"))+
  ylab("Percent with Sanders preference")+
  xlab("Age")+
  ggtitle("Model predictions for age vs. Sanders preference\nfor white voters, grouped by party")+
  theme(plot.title=element_text(hjust=.5))
```

```{r}
#plot data and prediction for non-white voters
ggp <- ggplot(age_bin_agg_all[age_bin_agg_all$race=="Non-White",], aes(x=agebin, y=sanders_preference,
                                     color=party, size=n))

ggp + geom_point()+
  geom_path(inherit.aes=F, data=data.frame(x=a, y=y_dem_nw),
             aes(x = a, y = y), color="blue")+
  geom_path(inherit.aes=F, data=data.frame(x=a, y=y_oth_nw),
             aes(x = a, y = y), color="green")+
    geom_path(inherit.aes=F, data=data.frame(x=a, y=y_rep_nw),
             aes(x = a, y = y), color="red")+
  #facet_grid(party~.)+
  scale_color_manual(values=c("blue", "green", "red"))+
  ylab("Percent with Sanders preference")+
  xlab("Age")+
  ggtitle("Model predictions for age vs. Sanders preference\nfor non-white voters, grouped by party")+
  theme(plot.title=element_text(hjust=.5))
```

The above plots show the original data with a bubble for every age group, with each color representing a different political party. For the sake of not having crowded graphs we've separated the white and non-white voters into two different graphs. 

Because our final model did not include interaction terms, all 6 lines - i.e. the models' predictions for each subgroup - are parallel. The model predicts the same age-preference relationship for all subgroups, and based on the subgroups boosts or lowers the probability (other > republican > democrat, and white > non-white)



# 3 Comment on Importance of Age and Client recommendation




The results of the model and the graphs can help inform our client on who to target for this marketing campaign. The model does suggest that younger voters are more likely to support Sanders, as each 10 year decrease in age corresponds to a 1.13 increase in the odds of supporting Sanders. 

The significant differences in supporting Sanders when looking at diferent party affiliations and races can help our client arrive at more targeted campaigns. Non-white Democrats appear to be the group least likely to support Sanders, as our model predicts them to have a less than 50% probability of supporting Sanders even at their youngest age range. (A 20 year old non-white Democrat is predicted to have a 41.0% chance of supporting Sanders). Non-white voters of other parties are also less likely to support Sanders than white voters, but our model does predict rates above 50% once you get below a certain age (Non-white Republicans age 39 and younger, as well as Non-white Other/Independents age 48 and younger are predicted to support Sanders with a probability greater than 50%)

Our client should attempt to target white voters who, across the sample, see an increase of 2.39 in odds of supporting Sanders compared to non-white voters. Our model predicts that white Republicans and Other/Independents have a greater than 50% chance of supporting Sanders across all age groups. For white Democrats, it appears that voters of age 61 and less are more than 50% likely to support Sanders.

All that being said, if our client is interested in targeting voters who support politically liberal candidates, these recommendations need to be taken with a grain of salt. Our model shows that Republicans are far more likely to prefer Sanders than Democrats. However, given the typical stance of Republican voters, this is likely more due to being against Clinton than supporting Sanders. It would likely be unwise to target Republicans in a campaign for liberal merchandise despite what these data suggest. Independent voters would likely be a good group to target. Our model predictions are able to provide ages within each subgroup where the chance of preferring Sanders rises above 50%, which could be natural cutoffs for a targeted marketing campaign.
