---
title: "lab5_samir"
author: "Samir Datta"
date: "December 10, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(reshape2)
library(plm)
load("driving.Rdata")
drivedata <- data
```
bac, perse, sbprim, sbsecon, sl70plus, gdl, perc14_24, unem, vehicmilespc


#Structure of the data

```{r}
head(drivedata)
unique(drivedata$year)
unique(drivedata$state)
```

The data is in long form. There are 25 rows per state, each representing a different year from 1980 to 2005. The data description data frame tells us that the "state" variable is a number that simply represents the 50 states in alphabetical order.

Some variables like the BAC and seatbelt law variable have more than two levels represented by multiple dummy coded variables. For example, BAC has three levels - no BAC law, BAC limit = .08, and BAC limit = .10. These three levels are coded in the two variables bac08 and bac10. Each of the years is also dummy coded in its own column.

An odd fluke in the data noticed above that there is no state number 2. The number is suppose to represent the states in alphabetical order, but 2 is skipped and the maximum number is 51 instead of 50. This will not affect our modeling in any way since the number is effectively a categorical variable and the values don't matter.


#EDA

##Total fatality rate
The primary variable of interest is "totfatrte", the total fatalities per 100,000 population.

```{r}
hist(drivedata$totfatrte, main="Histogram of total fatality per 100k")
```

The data has a slight positive skew. Note that this histogram is of all values of total fatality rate, across all states and years.


```{r}
ggplot(drivedata, aes(x=state, y=totfatrte, group=state)) + geom_boxplot()+
ylab("Total fatality rate")+xlab("State number")+ggtitle("Boxplot of total fatality rate per state")
```

Clearly, each state has different fatality rates, and some states simply have higher or lower rates regardless of the year or other variables.


```{r}
drivedata$statebucket <- cut(drivedata$state, 9)

ggplot(drivedata, aes(x=year, y=totfatrte, group=as.factor(state), color=as.factor(state%%9)))+geom_point()+
  geom_line()+facet_wrap(~as.factor(statebucket), scales="free")+
  ylab("Total Fatality Rate")+guides(fill=F)+
  ggtitle("Fatality rates from 1980-2005 for each state")
```


The above plot shows the fatality rates from 1980-205. Each set of dots connected with lines represents the data for a different state. 9 separate graphs are shown simply for the sake of avoiding clutter - the states have been binned by the number in the dataset (their alphabetical order).

This graph is simply to get an idea of the nature of year-to-year changes in fatlity rate. It seems for most states the rates steadily go down, but there are some interesting exceptions. It also seems that for a lot of states, the rates for the first one or two years is much higher, after which the rates drastically drop. The overall decrease in fatality rates over time is likely due to factors such as laws enacted.


##Laws
bac, perse, sbprim, sbsecon, sl70plus, gdl, perc14_24, unem, vehicmilespc

```{r}
hist(drivedata$bac08)
```
Many of the explanatory variables of interest (bac08, bac10, perse, sbprim, sbsecon, sl70plus, gdl) appear to be binary - 0 if the law was not in place, 1 if it was. As shown in the example histogram above, there are a very small number of fractional values. 


```{r}
head(drivedata[drivedata$year>1990,c("year", "bac08")], 10)
```

As shown in the example section of the data above, these fractional values occur between stretches of 0 and 1. There is no additional clarifying information in the dataset or data descriptions. As such, we are moving forward with the assumption that fractional values occur when the law was enacted part way through the year, and the value represents how much of the year the law was in place for.

Keeping these fractional values in our data will force the model to treat them as continuous variables, which is very problematic. We are not interested in the effect of a law if it is present for part of the year; rather, we want to know the binary effect of a law being in place. It is also very problematic to use a continuous variable in a linear model with such an odd and non-normal distribution, given the sparsity of the fractional values. As such, we are moving forward using the rounded values which will turn the fractional values into 0 or 1.


```{r}
#code to binarize all law variables
drivedata$bac08_binary <- ifelse(round(drivedata$bac08)==1, 1, 0)
drivedata$bac10_binary <- ifelse(round(drivedata$bac10)==1, 1, 0)
drivedata$perse_binary <- ifelse(round(drivedata$perse)==1, 1, 0)
drivedata$sbprim_binary <- ifelse(round(drivedata$sbprim)==1, 1, 0)
drivedata$sbsecon_binary <- ifelse(round(drivedata$sbsecon)==1, 1, 0)
drivedata$sl70plus_binary <- ifelse(round(drivedata$sl70plus)==1, 1, 0)
drivedata$gdl_binary <- ifelse(round(drivedata$gdl)==1, 1, 0)
drivedata$sl55_binary <- ifelse(round(drivedata$sl55)==1, 1, 0)
drivedata$sl65_binary <- ifelse(round(drivedata$sl65)==1, 1, 0)
drivedata$sl70_binary <- ifelse(round(drivedata$sl70)==1, 1, 0)
drivedata$sl75_binary <- ifelse(round(drivedata$sl75)==1, 1, 0)
drivedata$slnone_binary <- ifelse(round(drivedata$slnone)==1, 1, 0)
```

```{r}
year_law_agg <- with(drivedata, aggregate(cbind(bac08_binary, bac10_binary, perse_binary, sbprim_binary, sbsecon_binary, sl70plus_binary, gdl_binary), list(year=year), mean))
                     
year_law_melt <- melt(year_law_agg, id.vars="year")
                     
ggplot(year_law_melt, aes(x=year, y=value, color=variable, group=variable))+
  geom_line(size=2)+ylab("% of states with law in place")+xlab("Year")+
  ggtitle("Percent of states with law enacted from 1980-2005")+
  ylim(c(0,1))
```

This graph shows the % of states that had a certain law in place in every year.

The blood alcohol content laws are interesting. Clearly in the early 2000s there was a big movement to decrease the legal BAC from .10 to .08 given the opposite directions of those lines. A similar, but weaker, pattern is shown with the seatbelt law being primary or secondary around 1995. 

It is also interesting to note that both the graduated drivers license law and the speed limit being 70 or more changes did not happen at all until around 1995, and then fairly rapidly became more of a standard. Still, the speed limit change plateaus around 60%. It seems that about 60% of states adopted the speed limit change very quickly and the remaining states did not.


```{r}
drivedata$bac_combined <- ifelse(drivedata$bac08_binary==1, "bac08", 
                                 ifelse(drivedata$bac10_binary==1, "bac10", "noBAClaw"))
ggplot(drivedata, aes(x=bac_combined, y=totfatrte, color=bac_combined, group=bac_combined))+geom_boxplot()+
  ggtitle("Total fatality rate, grouped by BAC law")

```

Predictably, in states/years where the BAC limit was .08, there were lower fatality rates than when the limit was .10. States/years where there was no BAC law in place had the highest rates on average.

```{r}
drivedata$sb_combined <- ifelse(drivedata$sbprim_binary==1, "sbprim", 
                                 ifelse(drivedata$sbsecon_binary==1, "sbsecon", "nosblaw"))
ggplot(drivedata, aes(x=sb_combined, y=totfatrte, color=sb_combined, group=sb_combined))+geom_boxplot()+
  ggtitle("Total fatality rate, grouped by seatbelt law")

```

States/years without a seatbelt law have the highest fatality rate. States/years with the primary seatbelt law have lower rates than those with the secondary seatbelt law.

```{r}
drivedata$sl_combined <- ifelse(drivedata$sl55_binary, "sl55", 
                                 ifelse(drivedata$sl65_binary==1, "sl65", 
                                        ifelse(drivedata$sl70_binary==1, "sl70",
                                               ifelse(drivedata$sl75_binary==1, "sl75","slnone"))))
ggplot(drivedata, aes(x=sl_combined, y=totfatrte, color=sl_combined, group=sl_combined))+geom_boxplot()+
  ggtitle("Total fatality rate, grouped by Speed Limit Law")
```
States/years with lower speed limits have lower fatalaty rates with the exception of speed limit of 55.


```{r}
#bac, perse, sbprim, sbsecon, sl70plus, gdl, perc14_24, unem, vehicmilespc

ggplot(drivedata, aes(x=perse_binary, y=totfatrte, color=perse_binary, group=perse_binary))+geom_boxplot()+
  scale_x_continuous(breaks=c(0,1), labels=c("No law", "Law in place"))+
  ggtitle("Total fatality rate, grouped by administrative license revocation (per se law)")

ggplot(drivedata, aes(x=sl70plus_binary, y=totfatrte, color=sl70plus_binary, group=sl70plus_binary))+geom_boxplot()+
  scale_x_continuous(breaks=c(0,1), labels=c("No law", "Law in place"))+
  ggtitle("Total fatality rate, grouped by speed limit being 70+")

ggplot(drivedata, aes(x=gdl_binary, y=totfatrte, color=gdl_binary, group=gdl_binary))+geom_boxplot()+
  scale_x_continuous(breaks=c(0,1), labels=c("No law", "Law in place"))+
  ggtitle("Total fatality rate, grouped by graduated drivers license law")
```
  
  
  
The above boxplots show that fatality rates are lower when the "per se" law and the graduated drivers license law are in place, and higher when the speed limit is above 70. At least based on this plot, it seems the speed limit has the weakest effect. Raising the speed limit would intuitively increase fatality rates due to faster and more reckless driving. However, based on the previous time plot, the speed limit change appeared to happen around the same time as other safe laws were being enacted more often. Including all of these variables into a model should clarify their individual, ceteris paribus effects.
  
```{r}
ggplot(data, aes(x=minage, y=totfatrte))+
  geom_point()+geom_smooth(method="auto")+ggtitle("Total fatality rate, grouped by minimum age")
```
States/years with higher minimum driving age show lower faalaty rates

##Other explanatory variables

```{r}
par(mfrow=c(2,1))
hist(drivedata$perc14_24, main="Histogram of %population age 14-24")
boxplot(perc14_24~state, data=drivedata, main="%population age 14-24 by state")
```
The percent of population across states/years appears to be a normally distributed variable. This variable doesn't appear to be especially different from state to state - there are no states that tend to have an especially high or low % of ages 14-24.


```{r}
ggplot(drivedata, aes(x=perc14_24, y=totfatrte))+
  geom_point()+geom_smooth(method="loess")+ggtitle("%population age 14-24 vs. fatality rates")
```

There appears to be a linear relationship where states/years with a higher % of the population between ages 14-24 have higher fatality rates. Perhaps younger drivers are less safe.


```{r}
ggplot(drivedata, aes(x=year, y=perc14_24))+
  geom_point()+geom_smooth(method="loess")+ggtitle("%population age 14-24 from 1980-2005")
```

The % of the population between 14 and 24 very distinctly goes down from 1980 to about 1990, after which it flattens out to around 14%.


```{r}
par(mfrow=c(2,2))
hist(drivedata$unem, main="Histogram of unemployment")
hist(log(drivedata$unem), main="Histogram of unemployment (log-transformed)")
boxplot(log(unem)~state, data=drivedata, main="unemployment rate by state")
```

The unemployment rate, which is a percent, has a strong positive skew. The normality of the distribution improves when applying a log transformation.

Some states appear to have distinctly higher or lower unemployment rates.

```{r}
ggplot(drivedata, aes(x=log(unem), y=totfatrte))+
  geom_point()+geom_smooth(method="loess")+ggtitle("Unemployment rates vs. fatality rates")
```

There does not appear to be any discernible relationship between unemployment rates and driving fatality rates.


```{r}
ggplot(drivedata, aes(x=year, y=log(unem)))+
  geom_point()+geom_smooth(method="loess")+ggtitle("Unemployment rates per person vs. fatality rates")
```

Unemployment apears to have gone down over time, with a slight uptick in 1990.

```{r}
par(mfrow=c(2,2))
hist(drivedata$vehicmilespc, main="Histogram of vehicle miles per person")
boxplot(vehicmilespc~state, data=drivedata, main="vehicle miles per person by state")
plot(vehicmiles/statepop~vehicmilespc, data=drivedata, main="Vehicle miles/state population vs. 'vehicmilspc'")
```



Although there is no description given for how "vehicmilespc" was calculated, we have shown (in the third plot above) that it is the number of vehicle miles travelled (in billions) divided by the state population. Despite this attempt at correction, it is clear from the boxplots that some states tend to have a higher or lower value for this variable.


```{r}
ggplot(drivedata, aes(x=vehicmilespc, y=totfatrte))+
  geom_point()+geom_smooth(method="loess")+ggtitle("Vehicle miles per person vs. fatality rates")
```

Fatality rates appears to be positively correlated with vehicle miles, although these relationship may be driven by a relatively small number of data points with large influence.


```{r}
ggplot(drivedata, aes(x=year, y=vehicmilespc))+
  geom_point()+geom_smooth(method="loess")+ggtitle("Vehicle miles per person from 1980-2005")
```



The number of vehicle miles per person appears to steadily increase over time across states in a linear fashion.

2. How is the our dependent variable of interest *totfatrte* defined? What is the average of this variable in each of the years in the time period covered in this dataset? Estimate a very simple regression model of totfatrte on dummy variables for the years 1981 through 2004. What does this model explain? Describe what you find in this model. Did driving become safer over this period? Please provide a detailed explanation.

```{r}
ggplot(data, aes(factor(year), totfatrte)) +
  geom_boxplot() + 
  ggtitle("Fatalaty Rate by Year") + 
  theme(plot.title = element_text(lineheight=1, face="bold"),
        axis.text.x = element_text(size=10,angle=90)) 

plot(aggregate(drivedata$totfatrte, list(drivedata$year), mean), ylab = 'Average Fatalaty Rate', xlab = 'Year', main = 'Average Fatalaty Rate by Year')
```

*totfatrte* is defined as total fatalities per 100,000 population.  Fatalaty Rate trends down from 80s to the early 90s and is somewhat flat from later 90s to 2000s.  Steepest decline in the 80s.

```{r}
glm.mod <- glm(totfatrte~d81 + d82 + d83 + d84 + d85 + d86 + d87 + d88 + d89 + d90 + d91 + d92 + d93
               + d94 + d95 + d96 + d97 + d98 + d99 + d00 + d01 + d02 + d03 + d04, data = drivedata)

summary(glm.mod)
```

The model shows the average change is fatality rate by each year compared to 1980.  All years show a decrease in fatalaty rates when compared to 1980.  The only year where the change is not significant is 1981.  From the years 1982 to 2004 we see significant drops in fatalaty rates ranging from ~4.5% to ~8.8%.  It appears when comparing to 1980, from 1982 to 2004 driving did become safer when measured by fatalaty rates.

3. Expand your model in Exercise 2 by adding variables bac08, bac10, perse, sbprim, sbsecon, sl70plus, gdl, perc14_24, unem, vehicmilespc, and perhaps transformations of some or all of these variables. Please explain carefully your rationale, which should be based on your EDA, behind any transformation you made. If no transformation is made, explain why transformation is not needed. How are the variables bac8 and bac10 de???ned? Interpret the coe???cients on bac8 and bac10. Do per se laws have a negative e???ect on the fatality rate? What about having a primary seat belt law? (Note that if a law was enacted sometime within a year the fraction of the year is recorded in place of the zero-one indicator.) 

```{r}
glm.mod <- glm(totfatrte~d81 + d82 + d83 + d84 + d85 + d86 + d87 + d88 + d89 + d90 + d91 + d92 + d93
               + d94 + d95 + d96 + d97 + d98 + d99 + d00 + d01 + d02 + d03 + d04 + bac08_binary + bac10_binary + perse_binary + sb_combined + sl_combined + gdl_binary + perc14_24 + unem + vehicmilespc, data = drivedata)

summary(glm.mod)
```


**Please explain carefully your rationale, which should be based on your EDA, behind any transformation you made. If no transformation is made, explain why transformation is not needed. **

*TODO*

**How are the variables bac8 and bac10 de???ned?** 

BAC stands for Blood Alcohol Concentration, and BAC laws are legislations that punish driving with certain BAC. In this case, the BAC variables have more than two levels represented by multiple dummy coded variables. The variable three levels - no BAC law, BAC limit = .08, and BAC limit = .10. These three levels are coded in the two variables bac08 and bac10. 

**Interpret the coe???cients on bac8 and bac10. **

Both coefficients for bac08 and bac10 are highly statistically significant and the negative coefficients for both bac08 and bac10 suggest that BAC laws reduce the fatality rate under study. Curiously, the coefficient for bac08 for this model suggests that a limit for .08 has a more significant effect than .10, since the coefficient for .08 has a larger absolute value.

**Do per se laws have a negative e???ect on the fatality rate? What about having a primary seat belt law? (Note that if a law was enacted sometime within a year the fraction of the year is recorded in place of the zero-one indicator.) **

We can also observe that per se laws reduce the fatality rate (coefficient -7.740e-01, p-value < 0.001). Interestingly, our model doesn't support statistically significant support for the primary seat belt law coefficient.

**4. Reestimate the model from Exercise 3 using a ???xed e???ects (at the state level) model. How do the coe???cients on bac08, bac10, perse, and sbprim compare with the pooled OLS estimates? Which set of estimates do you think is more reliable? What assumptions are needed in each of these models? Are these assumptions reasonable in the current context?**

Let's first estimate a fixed effects model at the state level.

```{r}

plm.mod <- plm(totfatrte~d81 + d82 + d83 + d84 + d85 + d86 + d87 + d88 + d89 + d90 + d91 + d92 + d93
               + d94 + d95 + d96 + d97 + d98 + d99 + d00 + d01 + d02 + d03 + d04 + bac08_binary + bac10_binary + perse_binary + sb_combined + sl_combined + gdl_binary + perc14_24 + unem + vehicmilespc, data = plm.data(drivedata, c("state", "year")))

summary(plm.mod)
```

**How do the coe???cients on bac08, bac10, perse, and sbprim compare with the pooled OLS estimates? **

The BAC coefficients for the fixed effects model are relatively similar to those obtained in the pooled OLS model. In both models, BAC laws suggest a reduced fatality rate, while a BAC limit of .08 seems to have a larger effect on fatality rate than a BAC limit of .10. 

In the fixed effect model, the coefficient for per se laws is negative just as in the previous model, suggesting per se laws reduce the fatality rates. However, the coefficient for per se laws is one order of magnitude larger than in the pooled OLS model. Finally, while the pooled OLS model coefficient for seat belt primary laws was not statistically significant, the fixed effects model has a highly statistically significant model and a negative coefficient, suggesting that primary seat belt laws do in fact reduce fatality rates, which makes sense.


**Which set of estimates do you think is more reliable? What assumptions are needed in each of these models? Are these assumptions reasonable in the current context?**

The estimates from the fixed effects model should be more reliable.  Simple regression suffers more from omitted variable bias, specifically we need to assume that onobserved effects $a_i$ are not correlated with the explanatory variables.  If this is not true we have heterogeneity bias.  For the fixed effects estimator we difference away the time constant unobserved effects.  

First Difference:
$\Delta y_t = \delta_0 + \beta_1 \Delta x_i + \Delta u_i$

The key assumption for the first difference equation is that $\Delta u_i$ is not correlated with $\Delta x_i$.  This is a version of strict exogeneity.  If we violate this assumption we will have bias in the estimators.  We also need to assume that $\Delta x_i$ varies across states and time.  We do see that there is variation across explanatory across states and time intervals.

**5. Would you perfer to use a random effects model instead of the fixed effects model you build in *Exercise 4*? Why? Why not?**

**6. Suppose that *vehicmilespc*, the number of miles driven per capita, increases by 1,000. Using the FE estimates, what is the estimated effect on totfatrte? Be sure to interpret the estimate as if explaining to a layperson.**

Our coefficient for $vehicmilespc$ is 0.0009.  This means that there is a positive relationship between miles driven per capita and fatality rates, or the more miles driven the more likely you are the be in a fatal accident.  Specifically our estimator suggests that for every 1000 mile increase in miles per capita we expect a 0.9% increase in the fatality rate.

**7. If there is serial correlation or heteroskedasticity in the idiosyncratic errors of the model, what would be the consequences on the coefficient estimates and their standard errors?**

