---
title: 'Statistical Methods for Discrete Response, Time Series, and Panel Data (W271):
  Lab 3'
author: "Eric Yang, Samir Datta, Carlos Castro"
date: "October 25, 2017"
output:
  pdf_document: default
  html_document: default
---

# Instructions:

*  **Due Date: 11/17/2017 (by mid-night)**

* Submission:

    * Submit your own assignment via ISVC
    
    * Submit 2 files:
        1. A pdf file including the summary, the details of your analysis, and all the R codes used to produce the analysis. Please do not suppress the codes in your pdf file.
        
        2. R markdown file used to produce the pdf file
        
    * Each group only needs to submit one set of files
    
    * Use the following file naming convensation
    
        * SectionNumber_hw01_FirstNameLastNameFirstInitial.fileExtension
    
        * For example, if you are in Section 1 and have two students named John Smith and Jane Doe, you should name your file the following
            * Section1_hw01_JohnS_JaneD.Rmd
            * Section1_hw01_JohnS_JaneD.pdf
    
    * Although it sounds obvious, please write the name of each members of your group on page 1 of your report.
    
    * This lab can be completed in a group of up to 3 people. Each group only needs to make one submission. Although you can work by yourself, we encourage you to work in a group.

# Introduction and Objective

This lab is should be treated as a tutorial instead of a typical lab in this class. The objective of this "lab"" is to help you practice manipulating time series data in R using the *xts* time-series class.

Instead of having you read a bunch of documents before even having you develop any codes, I design a 3-step approach for you to walk through this lab: 

  1. Have you read only a couple of pages to get a very quick introduction and motivation of using the time-series class xts
  
  2. Have you developed some conduct to learn how to use xts to accomplish frequently-encountered tasks when working with time series data
  
  3. Have you gone back and studied the details behind the methods you used in your code development.

This lab/tutorial starts with a quick introduction to xts and zoo objects, followed by the concepts of creating an xts object and converting to an xts object from an imported dataset, explaining how to construct and deconstruct an xts object.

As in any data analysis, you most likely will have to combine dataset. We cover merging and modifying time series after studynig xts object construction. We cover different kinds of joins - outer, inner, left, and right join. 

I introduce the library ```quantmod``` and the ```getSymbols``` function to download the historical Twitter stock price directly from the Google website.  Unlike the unemployment time series, which comes in with monthly frequency, the Twitter stock price and volume series come in as the daily frequency, making them good candidates for learning how to merge time series of different time frequencies, an activity that you may have to implement a lot in practice. 

With time series of different time frequencies, inevitably one may have to fill in missing values.  Remember that the specific method for missing value imputation is context-dependent.  I cover a couple of methods but by no means endose them as the "to-go" methods for filling in missing values.

 Finally, this lab introduces two use techniques: differencing a time series (against itself) and apply various functions to time series, which is used frequently in rolling statistics calculation and time series aggregation.


# Materials Covered in this lab

  - Primarily the references listed in this document

  - Reference 
      - "xts: Extensible Time Series" by Jeffrey A. Ryan and Joshua M. Ulrich. 2008. (xts.pdf)
      - "xts FAQ" by xts Development Team. 2013 (xts_faq.pdf)
      - xts_cheatsheet.pdf

# Tasks 1:

  1. Read 
    A. the **Introduction** section (Section 1), which only has 1 page of reading of xts: Extensible Time Series" by Jeffrey A. Ryan and Joshua M. Ulrich
    B. the first three questions in"xts FAQ"
        a. What is xts?
        b. Why should I use xts rather than zoo or another time-series package?
        c. HowdoIinstallxts?
    C. The "A quick introduction to xts and zoo objects" section in this document
        
  2. Read the "A quick introduction to xts and zoo objects" of this document

# A quick introduction to xts and zoo objects

### xts
```xts```
  - stands for eXtensible Time Series
  - is an extended zoo object
  - is essentially matrix + (time-based) index (aka, observation + time)

  - xts is a constructor or a subclass that inherits behavior from parent (zoo); in fact, it extends the popular zoo class. As such. most zoo methods work for xts
  - is a matrix objects; subsets always preserve the matrix form
  - importantly, xts are indexed by a formal time object. Therefore, the data is time-stamped
  - The two most important arguments are ```x``` for the data and ```order.by``` for the index. ```x``` must be a vector or matrix. ```order.by``` is a vector of the same length or number of rows of ```x```; it must be a proper time or date object and be in an increasing order

# Task 2:

  1. Read 
    A. Section 3.1 of "xts: Extensible Time Series" by Jeffrey A. Ryan and Joshua M. Ulrich
    
    B. the following questions in "xts FAQ"
        a. How do I create an xts index with millisecond precision?
        b. OK, so now I have my millisecond series but I still can’t see the milliseconds displayed. What went wrong?

  2. Follow the following section of this document


# Creating an xts object and converting to an xts object from an imported dataset

We will create an xts object from a matrix and a time index. First, let's create a matrix and a time index.  The matrix, as it creates, is not associated with the time indext yet.

```{r}
# Create a matrix
x <- matrix(rnorm(200), ncol=2, nrow=100)
colnames(x) <- c("Series01", "Series02")
str(x)
head(x,10)

idx <- seq(as.Date("2015/1/1"), by = "day", length.out = 100)
str(idx)
head(idx)
tail(idx)
```

In a nutshell, xts is a matrix indexed by a time object. To create an xts object, we "bind" the object with the index.  Since we have already created a matrix and a time index (of the same length as the number of rows of the matrix), we are ready to "bind" them together. We will name it *X*.

```{r}
library(xts)
X <- xts(x, order.by=idx)
str(X)
head(X,10)
```
As you can see from the structure of an ```xts``` objevct, it contains both a data component and an index, indexed by an objevct of class ```Date```.

**xtx constructor**
```
xts(x=Null,
    order.by=index(x),
    frequency=NULL,
    unique=NULL,
    tzone=Sys.getenv("TZ"))
```
As mentioned previous, the two most important arguments are ```x``` and ```order.by```.  In fact, we only use these two arguments to create a xts object before.


With a xts object, one can decompose it.

### Deconstructing xts
```coredata()``` is used to extract the data component
```{r}
head(coredata(X),5)
```

```index()``` is used to extract the index (aka times)
```{r}
head(index(X),5)
```
  
### Conversion to xts from other time-series objects

We will use the same dataset "bls_unemployment.csv" that we used in the last live session to illustarte the functions below.


```{r}
# Set working directory
wd <-"~/Documents/Teach/Cal/w271/Labs/Lab3"
setwd(wd)

# Clean up the workspace before we begin
rm(list = ls())

df <- read.csv("bls_unemployment.csv", header=TRUE, stringsAsFactors = FALSE)

# Examine the data structure
  str(df)
  names(df)
  head(df)
  tail(df)

#table(df$Series.id, useNA = "always")
#table(df$Period, useNA = "always")

# Convert a column of the data frame into a time-series object
unemp <- ts(df$Value, start = c(2007,1), end = c(2017,1), frequency = 12)
  str(unemp)
  head(cbind(time(unemp), unemp),5)

# Now, let's convert it to an xts object
df_matrix <- as.matrix(df)
  head(df_matrix)
  str(df_matrix)
  rownames(df)

unemp_idx <- seq(as.Date("2007/1/1"), by = "month", length.out = 
length(df[,1]))
  head(unemp_idx)

unemp_xts <- xts(df$Value, order.by = unemp_idx)
  str(unemp_xts)
  head(unemp_xts)
```

# Task 3:

  1. Read 
    A. Section 3.2 of "xts: Extensible Time Series" by Jeffrey A. Ryan and Joshua M. Ulrich
    
  2. Follow the following section of this document
  
# Merging and modifying time series

One of the key strengths of ```xts``` is that it is easy to join data by column and row using a only few different functions. It makes creating time series datasets almost effortless.

The important criterion is that the xts objects must be of identical type (e.g. integer + integer), or be POSIXct dates vector, or be atomic vectors of the same type (e.g. numeric), or be a single NA. It does not work on data.frames with various column types.

The major functions is ```merge```.  It works like ```cbind``` or SQL's ```join```:

Let's look at an example. It assumes that you are familiar with concepts of inner join, outer join, left join, and right join.

```{r}
library(quantmod)
getSymbols("TWTR", src="google")
head(TWTR)
str(TWTR)
```

Note that the date obtained from the getSymbols function of teh quantmod library is already an xts object.  As such, we can merge it directly with our unemployment rate xts object constructed above. Nevertheless, it is instructive to examine the data using the View() function to ensure that you understand the number of observations resulting from the joined series.

```{r}
# 1. Inner join
TWTR_unemp01 <- merge(unemp_xts, TWTR, join = "inner")
  str(TWTR_unemp01)
  head(TWTR_unemp01)

# 2. Outer join (filling the missing observations with 99999)
# Basic argument use
TWTR_unemp02 <- merge(unemp_xts, TWTR, join = "outer", fill = 99999)
  str(TWTR_unemp02)
  head(TWTR_unemp02)
  #View(TWTR_unemp02)

# Left join
TWTR_unemp03 <- merge(unemp_xts, TWTR, join = "left", fill = 99999)
  str(TWTR_unemp03)
  head(TWTR_unemp03)
  #View(TWTR_unemp03)
  
# Right join
TWTR_unemp04 <- merge(unemp_xts, TWTR, join = "right", fill = 99999)
  str(TWTR_unemp04)
  head(TWTR_unemp04)
  #View(TWTR_unemp04)
```

# Missing value imputation
xts also offers methods that allows filling missing values using last or previous observation. Note that I include this simply to point out that this is possible. I by no mean certify that this is the preferred method of imputing missing values in a time series.  As I mentioned in live session, the specific method to use in missing value imputation is completely context dependent.

Filling missing values from the last observation
```{r}
# First, let's replace the "99999" values with NA and then exammine the series. 

# Let's examine the first few dozen observations with NA
TWTR_unemp02['2013-10-01/2013-12-15'][,1]

# Replace observations with "99999" with NA and store in a new series
unemp01 <- TWTR_unemp02[, 1]
unemp01['2013-10-01/2013-12-15']
str(unemp01)
head(unemp01)
#TWTR_unemp02[, 1][TWTR_unemp02[, 1] >= 99990] <- NA

unemp02 <- unemp01
unemp02[unemp02 >= 99990] <- NA

cbind(unemp01['2013-10-01/2013-12-15'], unemp02['2013-10-01/2013-12-15'])

# Impute the missing values (stored as NA) with the last observation
#TWTR_unemp02_v2a <- na.locf(TWTR_unemp02[,1], 
#                            na.rm = TRUE, fromLast = TRUE) 
unemp03 <- unemp02
unemp03 <- na.locf(unemp03, na.rm = TRUE, fromLast = FALSE) 

# Examine the pre- and post-imputed series
#cbind(TWTR_unemp02['2013-10-01/2013-12-30'][,1], TWTR_unemp02_v2a['2013-10-01/2013-12-15'])
cbind(unemp01['2013-10-01/2013-12-15'], unemp02['2013-10-01/2013-12-15'],
unemp03['2013-10-01/2013-12-15'])
```

Another missing value imputation method is linear interpolation, which can also be easily done in xts objects. In the following example, we use linear interpolation to fill in the NA in between months.  The result is stored in ```unemp04```. Note in the following the different ways of imputing missing values.

```{r}
unemp04 <- unemp02
#unemp04['2013-10-01/2014-02-01']
unemp04 <- na.approx(unemp04, maxgap=31)
#unemp04['2013-10-01/2014-02-01']

round(cbind(unemp01['2013-10-01/2013-12-15'], unemp02['2013-10-01/2013-12-15'],
unemp03['2013-10-01/2013-12-15'],
unemp04['2013-10-01/2013-12-15']),2)
```

## Calculate difference in time series
A very common operation on time series is to take a difference of the series to transform a non-stationary serier to a stationary series. First order differencing takes the form $x(t) - x(t-k)$ where $k$ denotes the number of time lags. Higher order differences are simply the reapplication of a difference to each prior result (like a second derivative or a difference of the difference).

Let's use the ```unemp_xts``` series as examples:
```{r}
str(unemp_xts)
unemp_xts
 
diff(unemp_xts, lag = 1, difference = 1, log = FALSE, na.pad = TRUE)

# calculate the first difference of AirPass using lag and subtraction
#AirPass - lag(AirPass, k = 1)

# calculate the first order 12-month difference if AirPass
diff(unemp_xts, lag = 12, differences = 1)
```

# Task 4:

  1. Read 
    A. Section 3.4 of "xts: Extensible Time Series" by Jeffrey A. Ryan and Joshua M. Ulrich
    
    B. the following questions in "xts FAQ"
        a. I am using apply() to run a custom function on my xts series. Why the returned matrix has di↵erent dimensions than the original one?

  2. Follow the following two sections of this document

# Apply various functions to time series

The family of ```apply``` functions perhaps is one of the most powerful R function families. In time series, ```xts``` provides ```period.apply```, which takes (1) a time series, (2) an index of endpoints, and (3) a function to apply.  It takes the following general form:
```
period.apply(x, INDEX, FUN, ...)
```

As an example, we use the Twitter stock price series (to be precise, the daily closing price), create an index storing the points corresopnding to the weeks of the daily series, and apply functions to calculate the weekly mean. 

```{r}
# Step 1: Identify the endpoints; in this case, we use weekly time interval. That is, we extract the end index on each week of the series

#View(TWTR)
head(TWTR)
TWTR_ep <- endpoints(TWTR[,4], on = "weeks")
#TWTR_ep

# Step 2: Calculate the weekly mean
TWTR.Close_weeklyMean <- period.apply(TWTR[, 4], INDEX = TWTR_ep, FUN = mean)
head(round(TWTR.Close_weeklyMean,2),8)
```

The power of the apply function really comes with the use of custom-defined function. For instance, we can easily 

```{r}
f <- function(x) {
  mean <- mean(x)
  quantile <- quantile(x,c(0.05,0.25,0.50,0.75,0.95))
  sd <- sd(x)
  
  result <- c(mean, sd, quantile)
  return(result)
}
head(round(period.apply(TWTR[, 4], INDEX = TWTR_ep, FUN = f),2),10)
```

# Calculate basic rolling statistics  of series by month
Using ```rollapply```, one can calculate rolling statistics of a series:

```{r}
# Calculate rolling mean over a 10-day period and print it with the original series
head(cbind(TWTR[,4], rollapply(TWTR[, 4], 10, FUN = mean, na.rm = TRUE)),15)
```

# Task 5:
1. Read AMAZ.csv and UMCSENT.csv into R as R DataFrames

## 5.1 Answer

```{r}
library(xts)
amazon.data = data.frame(read.csv('AMAZ.csv'))
umcsent.data = data.frame(read.csv('UMCSENT.csv'))
```

2. Convert them to xts objects

## 5.2 Answer

```{r}

amazon.xts = xts(amazon.data[,2:6], order.by = as.Date(amazon.data$Index))
umcsent.xts = xts(umcsent.data[,2], order.by = as.Date(umcsent.data$Index))

head(amazon.xts)
head(umcsent.xts)

```

3. Merge the two set of series together, perserving all of the obserbvations in both set of series
    a. fill all of the missing values of the UMCSENT series with -9999

## 5.3 Answer

### 5.3.a Answer

```{r}
# First merge the series
UMCSENT01 = merge(amazon.xts, umcsent.xts, join = "outer")

# We are asked to fill the missing values of UMCSENT
UMCSENT01$umcsent.xts[is.na(UMCSENT01$umcsent.xts)] <- -9999

# Display a summary
summary(UMCSENT01)
```


    b. then create a new series, named UMCSENT02, from the original  UMCSENT series replace all of the -9999 with NAs

### 5.3.b Answer

```{r}
# Transform -9999 records in umcsent.xts to NA
UMCSENT02 = UMCSENT01
UMCSENT02$umcsent.xts[UMCSENT02$umcsent.xts <= -9999] <- NA

# Display a summary
summary(UMCSENT02)
```

    c. then create a new series, named UMCSENT03, and replace the NAs with the last observation

### 5.3.c Answer

```{r}
# na.locf is designed exactly for this: https://www.rdocumentation.org/packages/zoo/versions/1.8-0/topics/na.locf
# "Generic function for replacing each NA with the most recent non-NA prior to it."
UMCSENT03 = UMCSENT02
UMCSENT03$umcsent.xts <- na.locf(UMCSENT03$umcsent.xts, fromLast = FALSE, na.rm = TRUE) 

# Display a summary
summary(UMCSENT03)
```
    d. then create a new series, named UMCSENT04, and replace the NAs using linear interpolation.

### 5.3.d Answer

```{r}
# na.approx does exactly what is asked: https://www.rdocumentation.org/packages/zoo/versions/1.8-0/topics/na.approx
UMCSENT04 = UMCSENT02
UMCSENT04$umcsent.xts <- na.approx(UMCSENT04$umcsent.xts)

# Display a summary
summary(UMCSENT04)
```
    e. Print out some observations to ensure that your merge as well as the missing value imputation are done correctly. I leave it up to you to decide exactly how many observations to print; do something that makes sense. (Hint: Do not print out the entire dataset!)

### 5.3.e Answer

Lets first get the number of observations:

```{r}
length(UMCSENT01[,1])
```
There are 1619 observations. However, printing the first results does not make sense, since the umcsent series starts before than the amazon series.

We choose to display the latest records where amazon observations are available.

#### 5.3.a display

```{r}
tail(UMCSENT01[!is.na(UMCSENT01$AMAZ.Open)])
```

#### 5.3.b display

```{r}
tail(UMCSENT02[!is.na(UMCSENT02$AMAZ.Open)])
```

#### 5.3.c display

```{r}
tail(UMCSENT03[!is.na(UMCSENT03$AMAZ.Open)])
```

#### 5.3.d display

```{r}
tail(UMCSENT04[!is.na(UMCSENT04$AMAZ.Open)])
```

4. Calculate the daily return of the Amazon closing price (AMAZ.close), where daily return is defined as $(x(t)-x(t-1))/x(t-1)$. Plot the daily return series.

```{r}
head(amazon.data)

AMAZ.Returns = c(NA, diff(amazon.data$AMAZ.Close, lag = 1, difference = 1, log = FALSE, na.pad = TRUE)) / amazon.data$AMAZ.Close

amazon.data = cbind(amazon.data, AMAZ.Returns)
amazon.xts = xts(amazon.data[,2:7], order.by = as.Date(amazon.data$Index))

# TODO: GGPLOT + legend + title
plot(amazon.xts$AMAZ.Returns, type = 'l')
```

5. Create a 20-day and a 50-day rolling mean series from the AMAZ.close series.

```{r}
amaz.close = cbind(amazon.xts[,4], rollapply(amazon.xts[, 4], 20, FUN = mean, na.rm = TRUE), rollapply(amazon.xts[, 4], 50, FUN = mean, na.rm = TRUE))
colnames(amaz.close) = c('ClosingPrice', 'TwentyDayRolling', 'FiftyDayRolling')
tail(amaz.close)

# TODO: GGPLOT + legend + title
plot(amaz.close$TwentyDayRolling)
lines(amaz.close$FiftyDayRolling, col = "green")
lines(amaz.close$ClosingPrice, col = "blue")

```
